---
bibliography: vignettes/using_the_gsna_package.bib
output: github_document
---
 
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

require( utils )

```

# GSNA version `r read.dcf( file = "DESCRIPTION" )[1,'Version']`

**Jonathan M. Urbach**

**Ragon Instute of MGH, MIT, and Harvard**  

**`r Sys.Date()`**  



<!-- badges: start -->
<!-- badges: end -->

GSNA stands for **G**ene **S**et **N**etwork **A**nalysis. GSNA is a toolkit for clustering gene sets based on metrics of similarity and distance such as the Jaccard and Szymkiewicz–Simpson overlap indices, and log Fisher *p*-values. The intended purpose of the GSNA package to provide a means to simplify data sets generated by pathways analysis methods such as **GSEA** (@subramanian_gene_2005,@mootha_pgc-1-responsive_2003), and **CERNO** (@zyla_gene_2019). GSNA can be used subsequent to pathways analysis methods that generate lists of gene sets with associated significance statistics, e.g. *p*-values. From such data, groups of similar pathways are inferred, greatly simplifying the task of analyzing complex pathways datasets. On the basis of similarity, networks and clusters (or subnets) can be generated and represented graphically, and statistical parameters can be assessed.


## Dependencies  

We recommend R version 4.0 and later for **GSNA**, though it may be installable on some later R version 3 distributions. In addition to base R, the **GSNA** package requires some other R packages including the following:

* circlize  
* DT  
* dendextend  
* dplyr  
* ggplot2  
* graphics  
* grDevices  
* igraph  
* Matrix  
* methods  
* psych  
* raster  
* stringr  
* stringi  
* stats  
* tibble  
* tidyr  
* tmod  
* utils  
* withr  
* Rcpp

Several of these packages have cascading dependencies, which become particularly important when installing from source. If binary R packages are available, such as for up-to-date Windows and Mac OS X R installations, we recommend installing those. Compiling these packages from source generally requires a C, C++, and/or Fortran compiler, which on Windows means installing Rtools for Windows, and for Mac OS X, Xcode. Linux installations generally include the required compilers, specifically the GCC compilers.    

**NOTE:** Particular care should be paid to the installation of the R **raster** package. The **raster** package requires the **terra** package, which in turn requires the **GEOS** C++ computational library, available from [https://libgeos.org/](https://libgeos.org/) to compile from source. Binary packages for the **GEOS** library are available for numerous Linux distributions, and may offer a more convenient alternative than installing **GEOS** from source.

## Installation  

We intend to make the **GSNA** package available on CRAN in the near future. If and when it is accepted by CRAN, you will be able to install **GSNA using the following command:

``` r
install.packages('GSNA')
```

In the meantime, if you have the **devtools** package installed, you can install the development version of GSNA directly from GitHub
like so:

``` r
devtools::install_github( repo = "https://github.com/JonathanUrbach/GSNA" )
```

Note: Currently, this method omits installation of the package's Roxygen2 man pages.  

If you have downloaded the source code from GitHub and opened the project in an Rstudio session, the following command can be used to install the package, including documentation:  

``` r
devtools::install( build_vignettes = TRUE, args = "--no-multiarch --with-keep.source" )
```


## Loading GSNA and Accessing Documentation    

To load GSNA, type the following in your R console:  

``` r 
library( GSNA )
```


To access a vignette containing additional information and usage examples, run:

``` r
vignette( "using_the_gsna_package" )
```

A package description and a list of function documentation can be obtained via:

``` r
help( package = "GSNA" )
```


# Citing GSNA

```{r echo=FALSE, results="asis"}
.cites <- readCitationFile( file = file.path( "inst", "CITATION" ) )
cat("To cite package 'GSNA' in publications, please use:\n\n"  )
print( .cites, bibtex = FALSE )
```



# References

<div id="refs"></div>


